Supplementary Material

### 2021-10-27 analysis/paper/paper.tex
preprint version

{\centering \includegraphics{/media/bjorn/hogwarts/Uni/publications/PhD/byocstarch/analysis/figures/protocol_overview}

@@

After 15 days, mineralisation was encouraged with a
calcium phosphate monofluorophosphate urea (CPMU) solution containing
20 mmol/l CaCl\textsubscript{2}, 12 mmol/l NaH\textsubscript{2}PO\textsubscript{4}, 5 mmol/l Na\textsubscript{2}PO\textsubscript{3}F, 500 mmol/l urea,
and (0.04 g/l MgCl)
(Pearce \& Sissons, 1987; Sissons et al., 1991).
The substrata were submerged in 1 ml/well CPMU for six minutes, five times
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated
for 10 days until the end of the experiment on day 24
(see \ref{fig:protocol-fig} for an overview of the protocol).

@@

An \(\alpha\)-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14.
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4°C until the assay was conducted on day 18.
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing
H\textsubscript{2}O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a
Multiskan FC Microplate Photometer (Thermo Scientific 51119000).
The protocol is a slightly modified version of an Enzymatic Assay of \(\alpha\)-Amylase
(\url{https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase}) (Bernfeld, 1955), which measures the amount of
maltose released from starch by \(\alpha\)-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.

@@

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 \(\tiny{M}\) ethylenediaminetetraacetic acid (EDTA)
(Le Moyne \& Crowther, 2021; Modi et al., 2020; Tromp et al., 2017),
and vortexing for 3 days until the sample was completely dissolved.
Twenty \(\mu\)l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using
the pipette to ensure that the counted transects were representative of the
whole slide. The count from the slide was extrapolated to the whole sample
(see Supplementary Material).

@@

Statistical analysis was conducted in R version 4.1.1 (2021-08-10) (R Core Team, 2020) and
the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
here (Müller, 2020), and patchwork (Pedersen, 2020).
A one-way ANOVA with sample weight as the dependent variable (DV) and treatment
as the grouping variable (GV) was conducted to explore the effect of the different
starch treatments on biofilm growth.

To calculate size proportions, mean counts for each treatment were taken across
both experimental plates for each treatment, resulting in a mean count for each
granule size category within each treatment.

@@

{\centering \includegraphics[width=0.5\linewidth]{/media/bjorn/hogwarts/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar}

@@

No \(\alpha\)-amylase activity was detected in any of the artificial
saliva samples from any of the days that were sampled. Only positive controls
contained amylase activity that could be detected in the assay, ranging from
3.4 to 10.3 U/mL enzyme
(Supplementary Material).
The results are not comparable to other studies presenting \(\alpha\)-amylase activity
levels in humans, as the unit definition may differ;
however, they are sufficient to show that there is no activity in the system.

\hypertarget{treatment-type-had-no-effect-on-biofilm-growth}{%
\subsection{Treatment type had no effect on biofilm growth}\label{treatment-type-had-no-effect-on-biofilm-growth}}

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
maximum length were underrepresented in the calculus samples compared to the original
starch solutions, an effect that was consistent across all three treatments,
while medium granules (10--20 \(\mu\)m) were often over-represented
(Table \ref{tab:sample-prop-tbl},
and
Figure \ref{fig:ratio-plots}).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 \(\mu\)m in maximum length, whereas wheat
granules generally only reach up to 35 \(\mu\)m
(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, pp. 174--176).
Granule morphology may also play a role. Large wheat granules
are lenticular and have a larger surface area compared
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area
(Jane et al., 1994; Reichert, 1913, pp. 364--365; Seidemann, 1966, pp. 174--176; van de Velde et al., 2002).
Another potentially important factor
from our results is the size of the calculus deposit. We found a
strong positive correlation between biofilm size and retained
starch granules (Figure \ref{fig:cor-plot}A);
a result that contradicts findings from archaeological contexts
(Dudgeon \& Tromp, 2014; Wesolowski et al., 2010).
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \ref{fig:cor-plot}B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches.
Further research is needed to determine why this differs from previous archaeological
findings.

@@

The mechanism by which starch granules are incorporated into plaque and calculus
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of considerable age
(Buckley et al., 2014; Henry et al., 2014; Wu et al., 2021).
Studies have also shown that not all starch granules come from a dietary source.
Other pathways include cross-contamination from plant interactions in soil, such
as palm phytoliths adhering to the skin of sweet potatoes
(Tromp \& Dudgeon, 2015),
or accidental ingestion not related to food consumption
(Radini et al., 2019; Radini et al., 2017).\\
When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are
only briefly present in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition through attachment to
tooth surfaces (including plaque and dental calculus) and mucous membranes
(Dodds \& Edgar, 1988; S. Kashket et al., 1991).
Bacteria also have the ability to adhere to starch granules
(Topping et al., 2003),
which would allow starches to attach to bacterial communities within the biofilm.
These granules are then
susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
by \(\alpha\)-amylase (S. Kashket et al., 1996).
The susceptibility of granules to hydrolysis depends on the crystallinity and size
of the starch granule, as well as the mode of processing. Smaller and pre-processed
(e.g., cooked) starch granules are more susceptible to enzymatic degradation,
while dehydrated starches will have a reduced susceptibility
(Björck et al., 1984; Franco et al., 1992; Haslam, 2004; Henry et al., 2009; Lingstrom et al., 1994).
Cracks on the surface of the dental calculus, as well as unmineralised islands
and channels may also be able to contain starch granules
(Charlier et al., 2010; Power et al., 2014; Tan, Gillam, et al., 2004).
Starch granules that are trapped in these pockets are (at least to some extent)
protected from aforementioned clearance mechanisms, especially once
a new layer of plaque has covered the surface of the plaque/calculus.
The size bias against large granules (\textgreater20 \(\mu\)m) from both wheat and potato
(Table \ref{tab:sample-prop-tbl}) may give further credence to
this incorporation pathway, as the smaller starch granules have an advantage over
larger granules, and can be stored in larger quantities.
This was also suggested by Power and colleagues (2014), who
observed clusters of starches within dental calculus, rather than an even
distribution across the surface of the dental calculus.
Granules trapped
in plaque/calculus may still be susceptible to hydrolysis, as \(\alpha\)-amylase has
the ability to bind to both tooth enamel and bacteria within a biofilm and retain
a portion of its hydrolytic activity
(Nikitkova et al., 2013; Scannapieco et al., 1993; Tan, Mordan, et al., 2004; Tan, Gillam, et al., 2004).
After the death of an individual, starches within dental calculus are susceptible
to further degradation by post-depositional processes, depending on burial environment
(pH, temperature, moisture content, microorganisms)
(Franco et al., 1992; García-Granero, 2020; Haslam, 2004; Henry et al., 2009).
Future study should explore how burial affects the recovery of starch from the
biofilm model.

@@

All scripts and data used in the analysis are available on OSF
(\url{https://osf.io/uc5qy/}) and Github (\url{https://github.com/bbartholdy/byoc-starch}),
following the format provided by the rrtools package (Marwick, 2019).
More detailed protocols are available on OSF and protocols.io.

@@

\leavevmode\hypertarget{ref-bernfeldAmylase1955}{}%
Bernfeld, P. (1955). Amylases, α and β. In \emph{Methods in {Enzymology}} (Vol. 1, pp. 149--158). {Academic Press}. \url{https://doi.org/10.1016/0076-6879(55)01021-5}

@@

\leavevmode\hypertarget{ref-cummingsMayanCalculus1997}{}%
Cummings, L. S., \& Magennis, A. (1997). A phytolith and starch record of food and grit in {Mayan} human tooth tartar. In A. Pinilla, J. Juan-Tresserras, \& M. J. Machado (Eds.), \emph{The {State}-of-the-{Art} of {Phytoliths} in {Soils} and {Plants}}. {CSIC Press}. \url{http://books.google.com?id=j66CDVfVhwEC}

@@

\leavevmode\hypertarget{ref-R-rrtools}{}%
Marwick, B. (2019). \emph{Rrtools: Creates a reproducible research compendium} {[}Manual{]}. \url{https://github.com/benmarwick/rrtools}

@@

\leavevmode\hypertarget{ref-R-here}{}%
Müller, K. (2020). \emph{Here: A simpler way to find your files} {[}Manual{]}. \url{https://CRAN.R-project.org/package=here}

@@

\leavevmode\hypertarget{ref-nikitkovaStarchBiofilms2013}{}%
Nikitkova, A. E., Haase, E. M., \& Scannapieco, F. A. (2013). Taking the {Starch} out of {Oral Biofilm Formation}: Molecular {Basis} and {Functional Significance} of {Salivary} α-{Amylase Binding} to {Oral Streptococci}. \emph{Applied and Environmental Microbiology}, \emph{79}(2), 416--423. \url{https://doi.org/10.1128/AEM.02581-12}

@@

\leavevmode\hypertarget{ref-R-patchwork}{}%
Pedersen, T. L. (2020). \emph{Patchwork: The composer of plots} {[}Manual{]}. \url{https://CRAN.R-project.org/package=patchwork}

@@

\leavevmode\hypertarget{ref-R-base}{}%
R Core Team. (2020). \emph{R: A language and environment for statistical computing} {[}Manual{]}. {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing}. \url{https://www.R-project.org/}

@@

\leavevmode\hypertarget{ref-R-broom}{}%
Robinson, D., Hayes, A., \& Couch, S. (2021). \emph{Broom: Convert statistical objects into tidy tibbles} {[}Manual{]}. \url{https://CRAN.R-project.org/package=broom}

\leavevmode\hypertarget{ref-scannapiecoSalivaryAmylase1993}{}%
Scannapieco, F. A., Torres, G., \& Levine, M. J. (1993). Salivary α-amylase: Role in dental plaque and caries formation. \emph{Critical Reviews in Oral Biology \& Medicine}, \emph{4}(3), 301--307.

>>>>

{\centering \includegraphics{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/protocol_overview}

@@

After 15 days, mineralisation was encouraged with a
calcium phosphate monofluorophosphate urea (CPMU) solution containing
20 mmol/l CaCl\textsubscript{2}, 12 mmol/l NaH\textsubscript{2}PO\textsubscript{4}, 5 mmol/l Na\textsubscript{2}PO\textsubscript{3}F, 500 mmol/l urea,
and (0.04 g/l MgCl)
(Pearce \& Sissons, 1987; Sissons et al., 1991).
The substrata were submerged in 1 ml/well CPMU for six minutes, five times
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated
for 10 days until the end of the experiment on day 24
(see Figure \ref{fig:protocol-fig} for an overview of the protocol).

@@

An \(\alpha\)-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14.
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4°C until the assay was conducted on day 18.
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing
H\textsubscript{2}O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a
Multiskan FC Microplate Photometer (Thermo Scientific 51119000).
The protocol is a slightly modified version of an Enzymatic Assay of \(\alpha\)-Amylase
(\url{https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase}) (Bernfeld, 1955), which measures the amount of
maltose released from starch by \(\alpha\)-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose in 6 minutes.

@@

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 \(\tiny{M}\) ethylenediaminetetraacetic acid (EDTA)
(Le Moyne \& Crowther, 2021; Modi et al., 2020; Tromp et al., 2017),
and vortexing for 3 days until the sample was completely dissolved.
Twenty \(\mu\)l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using
the pipette to ensure that the counted transects were representative of the
whole slide. The count from the slide was extrapolated to the whole sample
(see Supplementary Material for more detail).

@@

Statistical analysis was conducted in R version 4.1.1 (2021-08-10) (R Core Team, 2020) and
the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
here (Müller, 2020), and patchwork (Pedersen, 2020).

To see if biofilm growth was differently affected by starch treatments, a
one-way ANOVA with sample weight as the dependent variable (DV) and treatment
as the grouping variable (GV) was conducted.
To analyse granule counts and calculate size proportions, mean counts for each
treatment were taken across
both experimental plates, resulting in a mean count for each
granule size category within each treatment.

@@

{\centering \includegraphics[width=0.5\linewidth]{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar}

@@

No \(\alpha\)-amylase activity was detected in any of the artificial
saliva samples from any of the days that were sampled. Only positive controls
contained amylase activity that could be detected in the assay, ranging from
3.4 to 10.3 U/mL enzyme
(full results can be found in the Supplementary Material).
The results are not comparable to other studies presenting \(\alpha\)-amylase activity
levels in humans, as the unit definition may differ;
however, they are sufficient to show that there is no activity in the system.

\hypertarget{treatment-type-had-minimal-effect-on-biofilm-growth}{%
\subsection{Treatment type had minimal effect on biofilm growth}\label{treatment-type-had-minimal-effect-on-biofilm-growth}}

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
maximum length were underrepresented in the calculus samples compared to the original
starch solutions, an effect that was consistent across all three treatments,
while medium granules (10--20 \(\mu\)m) were often over-represented
(Table \ref{tab:sample-prop-tbl},
and
Figure \ref{fig:ratio-plots}).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 \(\mu\)m in maximum length, whereas wheat
granules generally only reach up to 35 \(\mu\)m
(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, pp. 174--176).
Granule morphology may also play a role. Large wheat granules
are lenticular and have a larger surface area compared
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area
(Jane et al., 1994; Reichert, 1913, pp. 364--365; Seidemann, 1966, pp. 174--176; van de Velde et al., 2002).
Another potentially important factor
from our results is the size of the calculus deposit. We found a
strong positive correlation between biofilm size and retained
starch granules (Figure \ref{fig:cor-plot}A);
a result that contradicts findings from archaeological contexts
(Dudgeon \& Tromp, 2014; Wesolowski et al., 2010).
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \ref{fig:cor-plot}B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = 5.53 mg) were on average larger than mixed treatment samples (mean = 4.28 mg) (Table \ref{tab:anova-tbl}); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

@@

The mechanism by which starch granules are incorporated into plaque and calculus
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of considerable age
(Buckley et al., 2014; Henry et al., 2014; Wu et al., 2021).
Studies have also shown that not all starch granules come from a dietary source.
Other pathways include cross-contamination from plant interactions in soil, such
as palm phytoliths adhering to the skin of sweet potatoes
(Tromp \& Dudgeon, 2015),
or accidental ingestion not related to food consumption
(Radini et al., 2017; Radini et al., 2019).\\
When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are
only briefly present in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition through attachment to
tooth surfaces (including plaque and dental calculus) and mucous membranes
(Dodds \& Edgar, 1988; S. Kashket et al., 1991).
Bacteria also have the ability to adhere to starch granules
(Topping et al., 2003),
which would allow starches to attach to bacterial communities within the biofilm.
These granules are then
susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
by \(\alpha\)-amylase (S. Kashket et al., 1996).
The susceptibility of granules to hydrolysis depends on the crystallinity and size
of the starch granule, as well as the mode of processing. Smaller and pre-processed
(e.g., cooked) starch granules are more susceptible to enzymatic degradation,
while dehydrated starches will have a reduced susceptibility
(Björck et al., 1984; Franco et al., 1992; Haslam, 2004; Henry et al., 2009; Lingstrom et al., 1994).
Cracks on the surface of the dental calculus, as well as unmineralised islands
and channels may also be able to contain starch granules
(Charlier et al., 2010; Power et al., 2014; Tan, Gillam, et al., 2004).
Starch granules that are trapped in these pockets are (at least to some extent)
protected from aforementioned clearance mechanisms, especially once
a new layer of plaque has covered the surface of the plaque/calculus.
The size bias against large granules (\textgreater20 \(\mu\)m) from both wheat and potato
(Table \ref{tab:sample-prop-tbl}) may give further credence to
this incorporation pathway, as the smaller starch granules have an advantage over
larger granules, and can be stored in larger quantities.
This was also suggested by Power and colleagues (2014), who
observed clusters of starches within dental calculus, rather than an even
distribution across the surface of the dental calculus.
Granules trapped
in plaque/calculus may still be susceptible to hydrolysis, as \(\alpha\)-amylase has
the ability to bind to both tooth enamel and bacteria within a biofilm and retain
a portion of its hydrolytic activity
(Nikitkova et al., 2013; Scannapieco et al., 1993; Tan, Mordan, et al., 2004; Tan, Gillam, et al., 2004).
After the death of an individual, starches within dental calculus are susceptible
to further degradation by post-depositional processes, depending on burial environment
(pH, temperature, moisture content, microorganisms)
(Franco et al., 1992; García-Granero, 2020; Haslam, 2004; Henry et al., 2009).
Future study should explore how burial affects the recovery of starch from the
biofilm model.

@@

All scripts and data used in the analysis are available on OSF
(\url{https://osf.io/uc5qy/}) and Github (\url{https://github.com/bbartholdy/byoc-starch}),
following the format provided by the rrtools package (Marwick, 2019).
More detailed protocols are available on OSF (\url{https://osf.io/akevs/}) and protocols.io (\url{https://www.protocols.io/view/biofilm-growth-with-starch-treatment-bu7jnzkn}).
Additional tables and figures are available in the Supplementary Material.

@@

\leavevmode\hypertarget{ref-bernfeldAmylase1955}{}%
Bernfeld, P. (1955). Amylases, {\(\alpha\)} and {\(\beta\)}. In \emph{Methods in {Enzymology}} (Vol. 1, pp. 149--158). {Academic Press}. \url{https://doi.org/10.1016/0076-6879(55)01021-5}

@@

\leavevmode\hypertarget{ref-cummingsMayanCalculus1997}{}%
Cummings, L. S., \& Magennis, A. (1997). A phytolith and starch record of food and grit in {Mayan} human tooth tartar. In A. Pinilla, J. Juan-Tresserras, \& M. J. Machado (Eds.), \emph{The {State}-of-the-{Art} of {Phytoliths} in {Soils} and {Plants}}. {CSIC Press}.

@@

\leavevmode\hypertarget{ref-R-rrtools}{}%
Marwick, B. (2019). \emph{Rrtools: Creates a reproducible research compendium} {[}Manual{]}.

@@

\leavevmode\hypertarget{ref-R-here}{}%
Müller, K. (2020). \emph{Here: A simpler way to find your files} {[}Manual{]}.

@@

\leavevmode\hypertarget{ref-nikitkovaStarchBiofilms2013}{}%
Nikitkova, A. E., Haase, E. M., \& Scannapieco, F. A. (2013). Taking the {Starch} out of {Oral Biofilm Formation}: Molecular {Basis} and {Functional Significance} of {Salivary} {\(\alpha\)}-{Amylase Binding} to {Oral Streptococci}. \emph{Applied and Environmental Microbiology}, \emph{79}(2), 416--423. \url{https://doi.org/10.1128/AEM.02581-12}

@@

\leavevmode\hypertarget{ref-R-patchwork}{}%
Pedersen, T. L. (2020). \emph{Patchwork: The composer of plots} {[}Manual{]}.

@@

\leavevmode\hypertarget{ref-R-base}{}%
R Core Team. (2020). \emph{R: A language and environment for statistical computing} {[}Manual{]}. {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing}.

@@

\leavevmode\hypertarget{ref-R-broom}{}%
Robinson, D., Hayes, A., \& Couch, S. (2021). \emph{Broom: Convert statistical objects into tidy tibbles} {[}Manual{]}.

\leavevmode\hypertarget{ref-scannapiecoSalivaryAmylase1993}{}%
Scannapieco, F. A., Torres, G., \& Levine, M. J. (1993). Salivary {\(\alpha\)}-amylase: Role in dental plaque and caries formation. \emph{Critical Reviews in Oral Biology \& Medicine}, \emph{4}(3), 301--307.

### 2021-10-27 analysis/supplementary-materials/supp-mat.Rmd
preprint version

---
title: Supplementary Material
output: 
  html_document:
    keep_md: true
    df_print: paged
    toc: true
    toc_float: true
---

@@

@@

Extracted-granule counts separated by size:

```{r}
corr_counts_long %>%
  filter(size != "total",
         treatment != "control") %>%
  group_by(treatment, starch, size) %>%
  summarise(sd = sd(count, na.rm = T),
            count = mean(count, na.rm = T)) %>%
  #mutate(percent = count / sum(count, na.rm = T) * 100) %>% 
  ggplot(aes(x = starch, y = count, fill = size)) +
    geom_col(position = "dodge") +
    geom_errorbar(aes(ymin = count, ymax = count + sd), width = 0.2, position = position_dodge(0.9)) +
    facet_wrap(~ treatment, scales = "free") +
    scale_fill_viridis_d() +
    theme_bw()
```

Size distribution within the solutions (top) and samples (bottom):

```{r message=FALSE, warning=FALSE}
sol_size_pl <- sol_corr %>%
  group_by(solution, starch) %>%
  summarise(across(c(s, m, l, total), mean, na.rm = T)) %>%
  pivot_longer(cols = c(s,m,l, total), values_to = "count", names_to = "size") %>%
  filter(size != "total") %>%
  group_by(solution, starch) %>%
  mutate(percent = count / sum(count, na.rm = T) * 100) %>% 
  ggplot(aes(x = starch, y = percent, fill = size)) +
    geom_col(position = "dodge") +
    facet_wrap(~ solution, scales = "free_x") +
    scale_fill_viridis_d() +
    theme_bw() +
    labs(x = "")

@@

Separated correlation plots:

>>>>

---
title: "Supplementary Material"
output:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
    df_print: kable
---

@@

## Protocols

All protocols are available on [protocols.io](https://www.protocols.io).

Creating the artificial saliva: <https://www.protocols.io/view/artificial-saliva-bva9n2h6>

Creating the CPMU solution: <https://www.protocols.io/view/cpmu-bv8pn9vn>

Biofilm growth protocol: <https://www.protocols.io/view/biofilm-growth-with-starch-treatment-bu7jnzkn>

Amylase activity assay: <https://www.protocols.io/view/amylase-activity-bw8jphun>

@@

Extracted-granule counts separated by treatment and size, including error bars:

```{r counts-bar, fig.cap="l = large, m = medium, s = small."}
corr_counts_long %>%
  filter(size != "total",
         treatment != "control") %>%
  group_by(treatment, starch, size) %>%
  summarise(sd = sd(count, na.rm = T),
            count = mean(count, na.rm = T)) %>%
  #mutate(percent = count / sum(count, na.rm = T) * 100) %>% 
  ggplot(aes(x = starch, y = count, fill = size)) +
    geom_col(position = "dodge") +
    geom_errorbar(aes(ymin = count, ymax = count + sd), width = 0.2, position = position_dodge(0.9)) +
    facet_wrap(~ treatment, scales = "free") +
    scale_fill_viridis_d() +
    theme_bw()
```

Size distribution (in %) within the solutions (top) and samples (bottom):

```{r size-ratios-bar, fig.cap="l = large, m = medium, s = small.", message=FALSE, warning=FALSE}
sol_size_pl <- sol_corr %>%
  group_by(solution, starch) %>%
  summarise(across(c(s, m, l, total), mean, na.rm = T)) %>%
  pivot_longer(cols = c(s,m,l, total), values_to = "count", names_to = "size") %>%
  filter(size != "total") %>%
  group_by(solution, starch) %>%
  mutate(percent = count / sum(count, na.rm = T) * 100) %>% 
  ggplot(aes(x = starch, y = percent, fill = size)) +
    geom_col(position = "dodge") +
    facet_wrap(~ solution, scales = "free_x") +
    scale_fill_viridis_d() +
    theme_bw() +
    labs(x = "")

@@

Separated correlation plots. These are the same plots as in the main paper, just larger.
